Streaming $2$-Center with Outliers in High Dimension

### 2015-04-17 paper.tex
abstract + parallel + revision

\begin{abstract}

\end{abstract}

@@

\section{Introduction}
Metric $k$-center problem -- covering a set of point with $k$ congruent balls of minimum radius -- is a fundamental problem in computer science and has various applications in data mining and image processing.
One of $k$\nobreakdash-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set. In real world applications, due to commonness of noisy data, it is important to consider outlier points. Because even a small number of outliers ($\cO(k)$) can increase the radius of $k$\nobreakdash-center problem unboundedly.
In this paper we focus on streaming data model. In the streaming data model, input arrives as a sequence and there is not enough space to keep the whole data set. Algorithms in this model have limitations in space (typically sub-linear in input size) and processing time per item. This model is specially suitable for massive data sets, due the its sub-linear space complexity \cite{aggarwal2007data}.

In this paper our concern is about $2$-center problem in high dimensional Euclidean space.
The formal definition of problem is as follows. Given sequence $\mathcal{S}$ of points in $d$ dimensional Euclidean space (for arbitrarily large $d$), find two congruent balls of minimum radius, covering all but at most $z$ points.

$k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$ \cite{bern1996approximation}.
Gonzalez~\cite{gonzalez1985clustering} showed a $2$-approximation algorithm for general case of the problem.
For small $k$ and $d$, there are more efficient algorithms.
For Euclidean $1$-center in fixed dimensions, Chazelle~\etal~\cite{chazelle1996linear} gave a linear algorithm.
For $2$\nobreakdash-center in the plane the best known algorithm, given by Chan \cite{chan1999more}, runs in $\cO(n \log^2 n \log^2 \log n)$ and linear space.

@@

For the problem of $k$-center with $z$ outliers in streaming data model, McCutchen~\etal~\cite{mccutchen2008streaming} (using idea in Charikar~\etal \cite{charikar2001algorithms} as subroutine) give a $(3+\eps)$-approximation ($(4+\eps)$-approximation, depending on the used subroutine) algorithm using $\cO(\frac{zk}{\eps})$ space.
In special case of $k=1$ in fixed dimensions, there is a streaming $\cO(\frac{z}{\eps^{\cO(d)}})$-space algorithm \cite{agarwal2007space, zarrabi2011almost}.
For Euclidean spaces, Zarrabi-Zadeh \cite{zarrabi2009streaming} gave $(1.22 \times \sqrt{2})$\nobreakdash-approximation for $1$-center with $z$ outliers running in $\cO(dz^3 + (\frac{d}{\eps^3})\log(\frac{1}{\eps}))$ space and poly$(d)$ update time .

In this paper we first give a simple $2$-approximation algorithm for $1$-center problem with space complexity $\cO(z^2 + zd)$. We then show a $(1.8 + \eps)$-approximation streaming algorithm for $2$-center with $z$ outliers in Euclidean space, with space complexity polynomial in $z$ and $d$ and sub-linear in $n$.

@@

\section{A Simple $2$-Approximation Algorithm for $1$-Center with Outliers}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=17em]{figs/one-center}
	\caption{Output of Algorithm~\ref{alg:1center}}
	\label{fig:1center}
\end{figure}

% Behnam: Aks be nazaram gooya nist ziad. va inke jaii behesh refrence dade nashode!
% Kiana: hagh ba behname. akse bedune refrence nadarim asan. montaha man benazaram kheili ehtiaji nis behesh asan 
% Sahand: hazf konim ya pishnehadi baray e behboodesh darid?

@@

When $\card{B_2}$ reaches $(d+1)(z+1)$, the candidate point changes to center point of $B_2$ and $r_c$ is increased by a factor of $(2 + \frac{2}{C})$ to completly cover the new candidate point and it's surronding ball. It is easy to see that $(2 + \frac{2}{C})r_c < \frac{\delta}{2}$ provided $C \ge 14$ and thus by Observation~\ref{obs:intersection} the property is not violated.

@@

Let the new candidate point be the centerpoint of points in $B_2$, when $(z+1)(d+1)$ points have been added to $B_2$. We claim that with $B_2 = B(c_2, (2 + \frac{2}{C})r_c)$, ball $B(c_p, \frac{2\card{c_1c_p}}{C})$ completly covered.
If $c_p \in B_{2,2}^*$ then by Observation~\ref{obs:c+4}, $r^* \le \frac{\card{c_1c_p}}{C}$. Thus $B_{2,2}^* \subseteq B(c_p, \frac{2\card{c_1c_p}}{C})$.
Since:
\[
\frac{2\card{c_1c_p}}{C} \leq \frac{2}{C}(\card{c_1c_2} + r_c) \leq (1 + \frac{2}{C})r_c
\]
 then $B_{2,2}^* \subseteq B(c_p, (1 + \frac{2}{C})r_c) \subseteq B(c_2, (2 + \frac{2}{C})r_c)$. Thus supposing $C \ge 14$ and by Observation~\ref{obs:intersection}, $B_2 \cap B_{2,1}^* = \emptyset$. Thus the claim follows.

@@

{
\algrenewcommand\algorithmicindent{0.8em}%
\begin{algorithm}
\label{alg:case2}
\leavevmode
\begin{algorithmic}
\State $\delta \gets 0, r_c \gets 0$
\State $c_1 \gets p_1$
\For{$p \in P$}
	\If{at least $z+1$ points have been processed}
		\State $q \gets$  $(z+1)$-furthest point from $c_1$
	\Else
		\State $q \gets c_1$
	\EndIf
	\State $\delta \gets \frac{C}{C+4} \card{c_1q}$
	\If{!\Call{AddTo$B_1$}{$p$} and !\Call{AddTo$B_2$}{$p$}}
% Sahand: is ! a familiar notation for CS people? should we use \neg? or not?
		\State add $p$ to Buffer
		\While{$\card{\mbox{Buffer}} > z$}
			\If{$\card{B_2} \ge (d+1)(z+1)$}
				\State $B_1 \gets B_u$, $B_2 \gets \emptyset$
			\ElsIf{$c_2$ is set}
				\State $B_1 \gets B_1 \cup \set{c_2}$, $B_u \gets B_u \cup \set{c_2}$
			\EndIf
			\State $T \gets \mbox{Buffer} \cup B_2 - \set{c_2}$
			\State $\mbox{Buffer} \gets \emptyset, B_2 \gets \emptyset$
			\If{first iteration of while}
				\For{$p \in T$}
					\If{\Call{AddTo$B_1$}{$p$}}
						\State Remove $p$ from $T$.
					\EndIf
				\EndFor
			\EndIf
			\State $c_2 \gets$ arbitrary point in $T$
			\State $r_c \gets \frac{2}{C}\card{c_1 c_2}$
			\For{$p \in T$}
				\State \Call{AddTo$B_2$}{$p$}
			\EndFor				
			\State $\mbox{Buffer} \gets T \setminus B_2$
		\EndWhile
	\EndIf
\EndFor
\end{algorithmic}
\end{algorithm}
}

@@

\begin{theorem}
\label{thm:2nd-cmplx}
Space complexity of Algorithm~\ref{alg:case2} is $\cO(dz+S(n,z,d))$. It takes $\cO^*(dzT(n,z,d))$ time for update. Query time in Algorithm~\ref{alg:query} is $\cO(zQ(n,z,d) + dz(dz+zQ_o(zd,z,d)))$ and it returns two centers that guarantee a $(1.8 + \eps)$-approximation for $2$-center problem with outliers.
% Sahand: Our analysis requires that $\beta \le 1.8+\eps$. Should we mention that somewhere? Someone might suspect using a worse approximation algorithm for 1-center he can achieve the same result with better time complexity.
\end{theorem}

>>>>

\begin{abstract}
In this paper, we study the $2$-center streaming problem with $z$ outliers in a general space, which means finding two congruent balls covering all the points except $z$ outliers in a streaming work-space. Before considering this problem, we provide a  $2$-approximation algorithm for $1$-center with $z$ outliers in a general space. Our algorithm for the $2$-center problem uses an update time and working space polynomial in $d$ and sub-linear in $n$ to calculate a $(1.8+\eps)$-approximation solution.
\end{abstract}

@@

\section{Introduction}
Metric $k$-center problem -- covering a set of point with $k$ congruent balls of minimum radius -- is a fundamental problem in computer science and has various applications in data mining and image processing.
One of $k$\nobreakdash-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set. In real world applications, due to commonness of noisy data, it is important to consider outlier points. Because even a small number of outliers ($\cO(k)$) can increase the radius of $k$\nobreakdash-center problem unboundedly.
In this paper we focus on streaming data model. In the streaming data model, input data arrives as a sequence and there is not enough space to keep the whole data set. Algorithms in this model have limitations in space (typically sub-linear in input size) and processing time per item. This model is specially suitable for massive data sets, due to its sub-linear space complexity \cite{aggarwal2007data}.

The $k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$ \cite{bern1996approximation}.
Gonzalez~\cite{gonzalez1985clustering} showed a $2$-approximation algorithm for general case of the problem.
For small $k$ and $d$, there are more efficient algorithms.
For Euclidean $1$-center in fixed dimensions, Chazelle~\etal~\cite{chazelle1996linear} gave a linear algorithm.
For $2$\nobreakdash-center in the plane the best known algorithm, given by Chan \cite{chan1999more}, runs in $\cO(n \log^2 n \log^2 \log n)$ and linear space.

@@

For the problem of $k$-center with $z$ outliers in streaming data model, McCutchen~\etal~\cite{mccutchen2008streaming} (using the idea in Charikar~\etal \cite{charikar2001algorithms} as subroutine) give a $(3+\eps)$-approximation ($(4+\eps)$-approximation, depending on the used subroutine) algorithm using $\cO(\frac{zk}{\eps})$ space.
In special case of $k=1$ in fixed dimensions, there is a streaming $\cO(\frac{z}{\eps^{\cO(d)}})$-space algorithm \cite{agarwal2007space, zarrabi2011almost}.
For Euclidean spaces, Zarrabi-Zadeh \cite{zarrabi2009streaming} gave $(1.22 \times \sqrt{2})$\nobreakdash-approximation for $1$-center with $z$ outliers running in $\cO(dz^3 + (\frac{d}{\eps^3})\log(\frac{1}{\eps}))$ space and poly$(d)$ update time .

In this paper our concern is about $2$-center problem in high dimensional Euclidean space.
The formal definition of problem is as follows. Given sequence $\mathcal{S}$ of points in $d$ dimensional Euclidean space (for arbitrarily large $d$), find two congruent balls of minimum radius, covering all but at most $z$ points. In Section~\ref{sec:2apr}, we give a simple $2$-approximation algorithm for $1$-center problem with space complexity $\cO(z^2 + zd)$. In the next section, we show a $(1.8 + \eps)$-approximation streaming algorithm for $2$-center with $z$ outliers in Euclidean space, and its space complexity is polynomial in $z$ and $d$ and sub-linear in $n$. In Section~\ref{subsec:smaller} we run $z + 1$ parallel instances of the algorithm in order to find the first non-outlier point.
%Kiana: man ino neveshtam vali manzuretun in bud raje be parallel budan az in bishtar begim yani? yekam ziadi tekrar nemishe?

% Kiana: I think first we should mention what others have done and then start declaring our problem. That's why I changed the location of the previous paragraph.

@@

\section{A Simple $2$-Approximation Algorithm for $1$-Center with Outliers}
\label{sec:2apr}

@@

%
%\begin{figure}[h]
%	\centering
%	\includegraphics[width=17em]{figs/one-center}
%	\caption{Output of Algorithm~\ref{alg:1center}}
%	\label{fig:1center}
%\end{figure}

% Behnam: Aks be nazaram gooya nist ziad. va inke jaii behesh refrence dade nashode!
% Kiana: hagh ba behname. akse bedune refrence nadarim asan. montaha man benazaram kheili ehtiaji nis behesh asan 
% Sahand: hazf konim ya pishnehadi baray e behboodesh darid?
% Kiana: Man be nazaram vaghean chon kheili az idehaye basic ham baraye gheire streaming hamine va osulan idash yekam badihitowre hazf beshe. felan commentesh kardam age nazaretun chize digei bud avazesh konin.

@@

When $\card{B_2}$ reaches $(d+1)(z+1)$, the candidate point changes to center point of $B_2$ and $r_c$ is increased by a factor of $(2 + \frac{2}{C})$ to completely cover the new candidate point and it's surrounding ball. It is easy to see that $(2 + \frac{2}{C})r_c < \frac{\delta}{2}$ provided $C \ge 14$ and thus by Observation~\ref{obs:intersection} the property is not violated.

@@

Let the new candidate point be the centerpoint of points in $B_2$, when $(z+1)(d+1)$ points have been added to $B_2$. We claim that with $B_2 = B(c_2, (2 + \frac{2}{C})r_c)$, ball $B(c_p, \frac{2\card{c_1c_p}}{C})$ completely covered.
If $c_p \in B_{2,2}^*$ then by Observation~\ref{obs:c+4}, $r^* \le \frac{\card{c_1c_p}}{C}$. Thus $B_{2,2}^* \subseteq B(c_p, \frac{2\card{c_1c_p}}{C})$.
Since:
\[
\frac{2\card{c_1c_p}}{C} \leq \frac{2}{C}(\card{c_1c_2} + r_c) \leq (1 + \frac{2}{C})r_c
\]
 then $B_{2,2}^* \subseteq B(c_p, (1 + \frac{2}{C})r_c) \subseteq B(c_2, (2 + \frac{2}{C})r_c)$. Thus supposing $C \ge 14$ and by Observation~\ref{obs:intersection}, $B_2 \cap B_{2,1}^* = \emptyset$. Thus the claim follows.

@@

{
\algrenewcommand\algorithmicindent{0.6em}%
\begin{algorithm}
\label{alg:case2}
\leavevmode
\begin{algorithmic}
\State $\delta \gets 0, r_c \gets 0$
\State $c_1 \gets p_1$
\For{$p \in P$}
	\If{at least $z+1$ points have been processed}
		\State $q \gets$  $(z+1)$-furthest point from $c_1$
	\Else
		\State $q \gets c_1$
	\EndIf
	\State $\delta \gets \frac{C}{C+4} \card{c_1q}$
	\If{\Call{AddTo$B_1$}{$p$} = False and  \Call{AddTo$B_2$}{$p$} = False}
% Sahand: is ! a familiar notation for CS people? should we use \neg? or not?
% Kiana: Agree, What about this one?
		\State add $p$ to Buffer
		\While{$\card{\mbox{Buffer}} > z$}
			\If{$\card{B_2} \ge (d+1)(z+1)$}
				\State $B_1 \gets B_u$, $B_2 \gets \emptyset$
			\ElsIf{$c_2$ is set}
				\State $B_1 \gets B_1 \cup \set{c_2}$, $B_u \gets B_u \cup \set{c_2}$
			\EndIf
			\State $T \gets \mbox{Buffer} \cup B_2 - \set{c_2}$
			\State $\mbox{Buffer} \gets \emptyset, B_2 \gets \emptyset$
			\If{first iteration of while}
				\For{$p \in T$}
					\If{\Call{AddTo$B_1$}{$p$}}
						\State Remove $p$ from $T$.
					\EndIf
				\EndFor
			\EndIf
			\State $c_2 \gets$ arbitrary point in $T$
			\State $r_c \gets \frac{2}{C}\card{c_1 c_2}$
			\For{$p \in T$}
				\State \Call{AddTo$B_2$}{$p$}
			\EndFor				
			\State $\mbox{Buffer} \gets T \setminus B_2$
		\EndWhile
	\EndIf
\EndFor
\end{algorithmic}
\end{algorithm}
}

@@

\begin{theorem}
\label{thm:2nd-cmplx}
Space complexity of Algorithm~\ref{alg:case2} is $\cO(dz+S(n,z,d))$. It takes $\cO^*(dzT(n,z,d))$ time for update. Query time in Algorithm~\ref{alg:query} is $\cO(zQ(n,z,d) + dz(dz+zQ_o(zd,z,d)))$ and it returns two centers that guarantee a $(1.8 + \eps)$-approximation for $2$-center problem with outliers.
% Sahand: Our analysis requires that $\beta \le 1.8+\eps$. Should we mention that somewhere? Someone might suspect using a worse approximation algorithm for 1-center he can achieve the same result with better time complexity.
% Kiana: I don't think so.
\end{theorem}
